Free Probability Analysis for Reflective Surface aided Massive MIMO systems

### 2020-12-17 documentation/Optimization Work/10_17_2020.tex
Back at the paper. Just added a couple of revisions

\begin{document}
\maketitle
\tableofcontents
\section{Points from Last Meeting}
\begin{enumerate}
\item{High MSE Issue:} \par
	As discussed with was a normalization error. Now I normalize the transmit vector $\vs$ to have total power 1. I normalize the noise vector $\vn$ in the same way.
	Finally, for the linear beamformer $\mathbf{W}$, in the case of the matched filter, I normalize each column using $\frac{\mathbf{w}_i}{\|\mathbf{w}_i\|^2}$ so that
	$\Expect\left[\mH\mW\right] = \mI$.

	\begin{figure}[H]
	\includegraphics[width= 16cm,height = 13cm]{figures/Corrected_MSE}
	  \caption{MSE with/without optimization versus number of IRS elements with receiver noise.}
	  	  \label{MSE_opt_noise}
	\end{figure}

\item{Expectation Over Elements of Channel:} \par
For the MSE Objective function with deterministic $\mW $ is 
	\begin{equation}
	\Expect\left[\trace\left((\mathbf{s} - \mH \mW \mathbf{s} )  (\mathbf{s} - \mH \mW \mathbf{s} ) ^H  \right)\right]
	\end{equation}
		\begin{equation}
	\trace\left(\Expect\left[ (\mI - \mH \mW  )  (\mI - \mH \mW  ) ^H \right] \Expect\left[ \mathbf{s}\mathbf{s}^H\right]\right)
	\end{equation}
	Assuming equal power allocation here since matrix size are relatively large this becomes
	\begin{equation}
	\trace\left(\Expect\left[ (\mI - \mH \mW  )  (\mI - \mH \mW  ) ^H \right]\right).
	\end{equation}
	Because the beamformer, $\mathbf{W}$, is deterministic at this point, the optimization problem becomes
		\begin{subequations}
	\label{mse_problem_total}
	\begin{align}
	   	    \underset{\boldsymbol{\phi}}{\text{minimize }}
	    & \; - \trace \left( \Expect\left[\mathbf{H} \right] \mathbf{W} \right) - \trace \left( \Expect\left[\mathbf{H} \right]  \mathbf{W}\right)^* +
	    \trace\left(\Expect\left[\mathbf{H}^H \mathbf{H} \right] \mathbf{W} \mathbf{W}^H \right)
	     \\
	    \text{subject to  } \; &
	    | \phi_i | = 1   \; \forall i \in {1 \cdots	 K}
	\end{align}
	\end{subequations}	
	Assuming channel components corresponding to the IRS path to be zero-mean, this becomes 
		\begin{subequations}
	\label{mse_problem_total}
	\begin{align}
	   	    \underset{\boldsymbol{\phi}}{\text{minimize }}
	    & \; \trace\left(\Expect\left[\mathbf{H}^H \mathbf{H} \right] \mathbf{W} \mathbf{W}^H \right)
	     \\
	    \text{subject to  } \; &
	    | \phi_i | = 1   \; \forall i \in {1 \cdots	 K}
	\end{align}
	\end{subequations}	
	In general we assume the channels through the elements of the IRS are correlated and in this case we use the exponential model.
	Now we decompose the channel $\mH$ into the individual paths through the IRS  $\Sigma_{i=1}^{N_S} \phi_i \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T}^{\frac{1}{2}}$. Expanding the utility function in \ref{mse_problem_total} shows that all $\phi_i$ cancel out from any non-cross term. The resulting utility function only contains cross terms of the form 
	\begin{equation}
		\trace\left(\Expect\left[ \phi_i \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T}^{\frac{1}{2}} (\phi_i \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,j}\mathbf{h}^T_{1,j}
	\mR_{T}^{\frac{1}{2}})\right] \right)
	\end{equation}
	or equivalently
		\begin{equation}
		\phi_j ^* \phi_i  \trace\left(\mR_{R} \Expect\left[ \mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T} (\mathbf{h}_{2,j}\mathbf{h}^T_{1,j})^H\right] \right).
	\end{equation}
	Assuming the all elements of $\mathbf{h}_{2,i}$ and $\mathbf{h}^T_{1,i}$ to be independent, these cross terms will also have expected value of zero and 
	thus the final expression does not include any $\phi_i$.
\item{Deterministic Elements of Channel:} \par
	In the above problem, if the transmitter has access to the channel state information for the channels through the IRS, the optimization problem from the last report can be used with results seen in figure \ref{MSE_opt_noise}
\end{enumerate}
\end{document}

>>>>

\begin{document}
\maketitle
\tableofcontents
\section{Points from Last Meeting}
\begin{enumerate}
In the first iteration of CSI collection, 1 pilot is 
\item{Issue with Expectation Over Elements of Channel:} \par
	The MSE Objective function with deterministic $\mW $ is 
	\begin{equation}
	\Expect\left[\trace\left((\mathbf{s} - \mH \mW \mathbf{s} )  (\mathbf{s} - \mH \mW \mathbf{s} ) ^H  \right)\right]
	\end{equation}
		\begin{equation}
	\trace\left(\Expect\left[ (\mI - \mH \mW  )^H  (\mI - \mH \mW  ) \right] \Expect\left[ \mathbf{s}\mathbf{s}^H\right]\right)
	\end{equation}
	Assuming equal power allocation this becomes
	\begin{equation}
	\trace\left(\Expect\left[ (\mI - \mH \mW  )^H  (\mI - \mH \mW  )  \right]\right).
	\end{equation}
	Because the beamformer, $\mathbf{W}$, is deterministic, the optimization problem becomes
		\begin{subequations}
	\label{mse_problem_total}
	\begin{align}
	   	    \underset{\boldsymbol{\phi}}{\text{minimize }}
	    & \; - \trace \left( \Expect\left[\mathbf{H} \right] \mathbf{W} \right) - \trace \left( \Expect\left[\mathbf{H} \right]  \mathbf{W}\right)^* +
	    \trace\left(\Expect\left[\mathbf{H}^H \mathbf{H} \right] \mathbf{W} \mathbf{W}^H \right)
	     \\
	    \text{subject to  } \; &
	    | \phi_i | = 1   \; \forall i \in {1 \cdots	 K}
	\end{align}
	\end{subequations}	
	Assuming channel components corresponding to the IRS path to be zero-mean (reasonable), this becomes 
		\begin{subequations}
	\label{mse_problem_total}
	\begin{align}
	   	    \underset{\boldsymbol{\phi}}{\text{minimize }}
	    & \; \trace\left(\Expect\left[\mathbf{H}^H \mathbf{H} \right] \mathbf{W} \mathbf{W}^H \right)
	     \\
	    \text{subject to  } \; &
	    | \phi_i | = 1   \; \forall i \in {1 \cdots	 K}
	\end{align}
	\end{subequations}	
	In general we assume the channels through the elements of the IRS are correlated and in this case we use the exponential model.
	Now we decompose the channel $\mH$ into the individual paths through the IRS  $\Sigma_{i=1}^{N_S} \phi_i \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T}^{\frac{1}{2}}$. Expanding the utility function in \ref{mse_problem_total} shows that all $\phi_i$  from the channel $\mH$ cancel out from the non-cross terms. The resulting utility function only contains cross terms of the form 
	\begin{equation}
		\trace\left(\Expect\left[ \phi_i \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T}^{\frac{1}{2}} (\phi_j \mR_{R}^{\frac{1}{2}}\mathbf{h}_{2,j}\mathbf{h}^T_{1,j}
	\mR_{T}^{\frac{1}{2}})^H \right] \right)
	\end{equation}
	in which $i \neq j$, or equivalently
		\begin{equation}
		\phi_j ^* \phi_i  \trace\left(\mR_{R} \Expect\left[ \mathbf{h}_{2,i}\mathbf{h}^T_{1,i}
	\mR_{T} (\mathbf{h}_{2,j}\mathbf{h}^T_{1,j})^H\right] \right).
	\end{equation}
	Assuming all elements of the random vectors $\mathbf{h}_{2,i}$ and $\mathbf{h}^T_{1,i}$ to be independent (also for $i \neq j$) these cross terms will also have expected value of zero. As a result the final expression does not include any $\phi_i$ meaning that without having knowledge of the actual channels, the MSE cannot be optimized with respect to the phases.
\item{Deterministic Elements of Channel:} \par
	In the above problem, if the transmitter has access to the channel state information for the channels through the IRS, the optimization problem from the last report can be used with results seen in figure \ref{MSE_opt_noise}
\end{enumerate}
\end{document}

### 2020-12-17 documentation/Paper/abstract.tex
Back at the paper. Just added a couple of revisions

Surfaces which can control the phase of reflected radio waves, known as intelligent reflective surfaces (IRS), have been the subject of interest in recent years as a means to better control the wireless propagation environment. This work investigates the impact of an IRS on the MIMO communication systems. Using the tools of free probability, a system is analyzed in the asymptotic domain to derive an expression for the system capacity. A key finding is that the phase control offered by the IRS has no impact in the asymptotic domain. Motivated by the contradiction between this finding and the substantial previous literature investigating the optimization of the IRS, we consider the convergence of a realistic system towards asymptotic behavior. We then propose an optimization method for the non-convex problem of minimizing the Mean Square Error with respect to the IRS phases.

>>>>

Surfaces which can control the phase of reflected radio waves, known as intelligent reflective surfaces (IRS), have been the subject of interest in recent years as a means to better control the wireless propagation environment. This work investigates the impact of an IRS on MIMO (Multiple Input Multiple Output)  communication systems. Using the tools of free probability, a MIMO communication system is analyzed in the asymptotic domain to derive an expression for the system capacity. A key finding is that the phase control offered by the IRS has no impact on system capacity in the asymptotic domain. Motivated by the contradiction between this finding and the substantial previous literature investigating the optimization of the IRS, we consider the convergence of a realistic system towards the derived asymptotic behavior. We then propose an optimization method for the non-convex problem of minimizing the Mean Square Error with respect to the IRS phases.

### 2020-12-17 documentation/Paper/background_opening.tex
Back at the paper. Just added a couple of revisions

The spectrum of electromagnetic frequencies used by wireless communication has been expanded to progressively higher frequencies with new standards. Industry research for the 6G standard has already investigated systems using frequency bands in the 120, 200 and 340 GHz range \cite{Koziol}. This shift is motivated by large bands of currently unoccupied frequency and the possibility for decreased antenna size (allowing for increased antenna density) \cite{akyildiz2018combating}. One primary hurdle towards efficient communication at these frequencies is the high attenuation of radio energy due both to scattering effects and molecular absorption. \cite{TODO} show through extensive data collection that relevant attenuation will be (x). 
As a relevant example, radar systems have long used this band of frequencies due to the resulting high resolution. For example, one such system operating in the 2 - 4 GHz band requires 25 kW of power for a detection range of 400 km \cite{TODO}. 
Countering this type of loss is a key step towards meeting the increased data rates and reliability required for future networks. 
\par
In previous generations of wireless communications, attenuation due to path loss has been alleviated by reducing the effective transmission distance 
and supplementing the transmission power using relays \cite{dahlman20134g}. Two, key costs to deploying relays are the radio frequency chains used for receiving and transmitting, and the power used for transmission (CITE?). Deploying enough relays to counter
the attenuation levels relevant to new standards would pose impractical costs. One new solution that has attracted significant interest is the use of surfaces made of many, sub-wavelength sized elements with favorable electromagnetic reflection and whose reflective coefficient can be controlled to select the phase of the reflected wave. In this work, such a surface will generally be referred to as a reconfigurable intelligent surface (RIS).
While this work focuses on the theory of communications channels with such surfaces, it is worth noting that these surfaces have been successfully implemented in works such as \cite{tan2018enabling} in which an RIS with 224 elements was built and tested. 
\par
The motivating use for the RIS is to alleviate the power attenuation problem without incurring the costs of an rf-chain and transmission power 
associated with a relay. In addition, the absence of an rf-chain on the RIS offers some benefits over relays. First, reflection from an RIS does not introduce the noise associated with an rf-chain. Second, without the band-pass filters of an rf-chain, RISs have a frequency response over the entire spectrum (explain). Third, the RIS is inherently full-duplex, a feature that is difficult to implement for co-located rf-chains on an active relay. 
The thorough review in \cite{basar2019wireless} provides rigorous motivation of the use of RISs and a few notable points are included here.
Consider a single antenna, single receiver channel using an RIS with $N$ reflective elements. In this case, the phase control at the RIS can be used to add the multiple paths coherently at the receiver and intuitively increase the received signal power. While such results are motivating, there are two points that should be noted. First, in this example from \cite{basar2019wireless}, the received power would still scale with $N^2$ even if the reflective surfaces had no control over the reflection phase. Second, it is important to note that scaling of received SNR proportional to $N^2$ will only hold to the point limited by energy conservation \cite{bjornson2019demystifying}.
This leaves us with a question. What is the true potential benefits for the RIS in future systems?
Before investigating this question we review key aspects of MIMO systems need to fully develop a system model with which to explore the RIS.

>>>>

In new standards for wireless communication, the spectrum of electromagnetic frequencies used has been expanded to progressively higher frequencies. Industry research for the 6G standard has already investigated systems using frequency bands in the 120, 200 and 340 GHz range \cite{Koziol}. This shift is motivated by large bands of currently unoccupied frequency and the possibility for decreased antenna size (to enable increased antenna density) \cite{akyildiz2018combating}. One primary hurdle towards efficient communication at these frequencies is the high attenuation of radio energy due both to scattering effects and molecular absorption. \cite{TODO} show through extensive empirical results that relevant attenuation can reach (x). 
As a relevant example, radar systems have long used this band of frequencies due to the resulting high resolution. For example, one such system operating in the 2 - 4 GHz band requires 25 kW of power for a detection range of 400 km \cite{TODO} (NEED NEW EXAMPLE). 
Countering this power loss is a key step towards meeting the increased data rates and reliability required for future wireless communication networks. 
\par
In previous generations of wireless communications, attenuation due to path loss has been alleviated by incorporating relays into the system to reduce the point-to-point transmission distance and supplement the transmission power \cite{dahlman20134g}. Two key costs to deploying relays are the radio frequency chains used for receiving and transmitting, and the power used for transmission (CITE?). As of today, most relays are only capable of operating in a half-duplex mode and thus a third cost to the system (by reducing capacity) is often incurred by using relays. Deploying enough relays to counter
the attenuation levels relevant to new standards would pose impractical costs. One new solution that has attracted significant interest is the use of surfaces made of many, sub-wavelength sized elements with favorable reflection for the relevant electromagnetic spectrum and whose reflective coefficient can be controlled to select the phase of the reflected wave. In this work, such a surface will generally be referred to as an intelligent reflective surface (IRS).
While this work focuses on the theory of IRS communication channels, it is worth noting that these surfaces have been successfully implemented in works such as \cite{tan2018enabling} in which an IRS with 224 elements was built and tested. 
\par
The motivating use for the IRS is to alleviate the power attenuation problem without incurring the costs of an rf-chain and transmission power 
associated with an active relay. In addition, the absence of an rf-chain on the IRS offers some benefits over relays. First, reflection from an IRS does not introduce the noise associated with an rf-chain. Second, without the band-pass filters of an rf-chain, IRSs have a frequency response over the entire spectrum (explain). Third, the IRS is inherently full-duplex, a feature that is difficult to implement for co-located rf-chains on an active relay. 
The thorough review in \cite{basar2019wireless} provides rigorous motivation of the use of IRSs and a few notable points are included here.
Consider a single antenna, single receiver channel using an IRS with $N$ reflective elements. In this case, the phase control at the IRS can be used to add the multiple paths coherently at the receiver and intuitively increase the received signal power. While such results are motivating, there are two points that should be noted. First, in this example from \cite{basar2019wireless}, the received power would still scale with $N^2$ even if the reflective surfaces had no control over the reflection phase. Second, it is important to note that scaling of received SNR proportional to $N^2$ will only hold to the point limited by energy conservation \cite{bjornson2019demystifying}.
This leaves us with a question. What is the true potential benefits for the IRS in future systems?
Before investigating this question we review key aspects of MIMO systems need to fully develop a system model with which to explore the IRS.
